---
title: "Trp53 Baseline & activity"
output:
  html_document:
    number_sections: yes
    toc: yes
    #code_folding: hide
    toc_depth: 4
    toc_float:
      collapsed: false
    highlight: pygments
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_knit$set(root.dir = "~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/scMPRA/results/figures/figure3")

library(tidyverse)
library(patchwork)
library(Seurat)
library(viridis)

```

In this .Rmd, we reproduce panel d from figure 3. This involved two steps: 

1.    Compute cell type-specific baseline enhancer activities from random DNA background controls
2.    Resolve enhancer activity x sequence strength (affinity x number of motif repeats) relationship across cell types

# Baseline activity

```{r read in data , include=TRUE}

#reading in the integrated expression data 
integrated <- readRDS(file = "~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/integrated_seurat.rds")

#read in enhancer count data from Lib Alpha screen 
guides <- readRDS("~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libAlpha_matched_gfp_guide_enhancer_counts.rds")

```

Merge seurat object meta data with enhancer count table `guides`. 

```{r merge meta with guides}

#lets get the GRE counts 
meta_cells <- integrated@meta.data %>% tibble(cell = colnames(integrated)) 

guides <- guides |> 
  dplyr::mutate(cell = paste0("Trp53.", cell))
  
guides <- left_join(guides, meta_cells)

```

Define cell type colors. 

```{r cell type colors}

another_pal<- c("Eo/Baso/Mast" = "#C43E2F", 
                "Eo/Baso/Mast Progenitors" =  "#F26B55" , 
                "CD55+ immature" = "#E78C64", 
                "Meg/Ery" = "#EDB93E", 
                "Monocytes late" = "#6C9F8C",
                "Monocytes cycling" = "#217F5C",
                "Monocytes early" = "#33C38E",
                "Neutrophils" = "#25625E", 
                "Early Myelocytes" = "#A0F2E8", 
                "Monocyte precursor" = "#54BFB7", 
                "Unclear/Excluded" = "grey", 
                "Neutrophil precursor" = "#9ab1a9", 
                "CD55- immature" = "#d2f9f4", 
                "EMPP" = "#ff4dff",
                "LMPP" = "#9999ff",
                "HSC/MPP cycling" = "#7a0099", 
                "HSC/MPP" = "#9900cc")

```

Here, we are computing some more qc measure from the seurat object for our baseline estimation and in order to understand if baseline activity is correlated with some of these qc measures or overall cellular features. 

```{r qc measures}
#extract some qc metric from seurat object 
approach_a <- integrated@meta.data %>%
  filter(experiment == "Trp53") |> 
  group_by(SCT_snn_res.2) %>%
  summarise(
    rnaCount         = mean(nCount_RNA,      na.rm = TRUE),
    rnaCount.sd      = sd(nCount_RNA,        na.rm = TRUE),
    rnaFeature       = mean(nFeature_RNA,    na.rm = TRUE),
    rnaFeature.sd    = sd(nFeature_RNA,      na.rm = TRUE),
    .groups = "drop"
  )
```

Next, we aggreagate the enhancer activity per enhancer for each cluster separately at cluster resolution SCT_snn_res.2. 

```{r aggregate CRSlevel}

crs.level.summary <- guides %>%
  group_by(CRS, Seqname, TF.1, TF.2, TF1.affinity, TF2.affinity, nrepeats, spacer, ControlFor, sequence,
           TF1.orientation, TF2.orientation, design_rep, Library, ctrls, SCT_snn_res.2) %>%
  dplyr::summarise(
    ncells = n_distinct(cell),
    nguides = n_distinct(guide_id),
    mean_gfp = mean(gfp_count, na.rm = TRUE),
    median_gfp = median(gfp_count, na.rm = TRUE),
    mean_guide = mean(guide_count, na.rm = TRUE),
    median_guide = median(guide_count, na.rm = TRUE),
    sd_gfp = sd(gfp_count, na.rm = TRUE),
    ngfp = n_distinct(gfp_id),
    nFeatures = mean(nFeature_RNA), 
    nCount = mean(nCount_RNA)
  )

```

We merge enhancer-level activity summaries with cluster-level sequencing statistics and compute the mean GFP activity per enhancer and cluster. To stabilize variance and enable visualization of low activity values, GFP activity is log-transformed using a small pseudocount equal to half of the smallest non-zero mean GFP value.

```{r}
#join the qc/overall cell features with aggregated enhancer data 
crs.level.summary <- merge(crs.level.summary, approach_a)

#here we are scaling enhancer signal to relative cluster-wide GFP levels 
#crs.level.summary$norm.gfp <- crs.level.summary$mean_gfp / crs.level.summary$gfpCount

#this divides enhancer activity by the average number of detected RNA features 
crs.level.summary$norm.rnaFeatures <- crs.level.summary$mean_gfp / crs.level.summary$rnaFeature

#computes a pseudocount for the log transfomation, by finding the smalles positive GFP mean  
min.raw <- min(crs.level.summary$mean_gfp[crs.level.summary$mean_gfp>0])

#log transform enhancer activity 
crs.level.summary$logm = with(crs.level.summary, log(mean_gfp + min.raw/2))

#to obtain cell types for each cluster we are doing a majority vote 
formerge <- plyr::ddply(data.frame(SCT_snn_res.2 = integrated$SCT_snn_res.2, annotation = integrated$final_annotation), "SCT_snn_res.2",plyr::summarise, final_annotation = names(which.max(table(annotation))), n_cells = length(SCT_snn_res.2))

#and this we are merging with the crs.level.summary object 
crs.level.summary <- merge(crs.level.summary, formerge)

#filter cell types that are not in screen 
crs.level.summary <- crs.level.summary |> 
  filter(!final_annotation %in% c("Unclear/Excluded", "EMPP", "LMPP", "HSC/MPP", "HSC/MPP cycling")) 

```

Next step is to compute a baseline enhancer activity from the random DNA controls. 

```{r baseline computation}
#compute the baseline 
baseline <- crs.level.summary |> 
  filter(TF.1 == "inert") |> 
  group_by(SCT_snn_res.2) |> 
  summarise(baseline = median(mean_gfp, na.rm = TRUE), .groups = "drop") |> 
  left_join(approach_a, by = "SCT_snn_res.2")  |> 
  left_join(formerge,   by = "SCT_snn_res.2")

```

Next, we can plot if baseline enhancer activity is correlated with the average number of genes captured per cell (aggregated at the cluster level).  

```{r baseline against rna features }
ggplot(baseline, aes(x = baseline, y = rnaFeature, group = SCT_snn_res.2, color = final_annotation)) +
  geom_point() +
  scale_color_manual(values = another_pal)

```


Here we need to find a solution, this is not a good clustering. 

```{r}
baseline_summary <- baseline %>%
  left_join(formerge, by = intersect(names(baseline), names(formerge))) %>%  # see note below
  group_by(final_annotation) %>%
  summarise(
    value = log(mean(baseline + min.raw/2, na.rm = TRUE)),
    affinity = 0,
    .groups = "drop")

```


```{r plot baseline and celltype}

tp53 <- subset(crs.level.summary, TF.1 == "Trp53" & TF.2 == "")

baseline_plot <- ggplot(tp53, aes(x = TF1.affinity * nrepeats, y = logm, group = final_annotation , color = final_annotation)
) +
  geom_point(data = baseline_summary, aes(x = affinity, y = value), size = 3) + 
  geom_smooth(se = FALSE, span = 1, linewidth = 1.5) +
  scale_color_manual(values = another_pal) +
  theme_minimal() +
  theme(text=element_text(family="ArialMT"),  
        axis.ticks = element_blank(),
        panel.grid.major.y = element_blank()) + 
        #legend.position = "none") + 
  geom_vline(xintercept = c(0.5, 3.25, 6), linetype = 3, linewidth = 1) +
  labs(x = "Sequence Strength", y = "Log GFP/cell") 

baseline_plot

```

Save plot. 

```{r save plot, eval = FALSE}

setwd("~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/figure3/")
ggsave("./004_lars_plot_baseline.pdf", lars_plot, width = 10, height = 15, device = "pdf", units = "cm")

```

