---
title: "Trp53 Umaps & sequence-activity relationship"
output: html_document
date: "2026-03-13"
---

---
title: "Cell type-specificity of CREs"
output:
  html_document:
    number_sections: yes
    toc: yes
    #code_folding: hide
    toc_depth: 4
    toc_float:
      collapsed: false
    highlight: pygments
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_knit$set(root.dir = "~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/scMPRA/results/figures/figure2")

library(ROCR)
library(patchwork)
library(viridis)
library(tidyverse)
library(Seurat)

```

In this .Rmd, we reproduce panel b andc from figure 3. Where we show that Trp53 enhancer activity is increases linearily with sequence strengh of the synthetic enhancers, defined as affinity * number of repeats. 

First, we read in data. 

```{r load data}

#reading in the integrated expression data 
integrated <- readRDS(file = "~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/integrated_seurat.rds")

#read in enhancer count data from Lib Alpha screen 
guides <- readRDS("~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libAlpha_matched_gfp_guide_enhancer_counts.rds")

```

# UMAP Trp53 enhancers 

First, we merge the umap coordinates with the enhancer data. 

```{r extract umap, include=TRUE}

#get the umap coord & join it with enhancer count matrix  
umapCoord <- as.data.frame(Embeddings(object = integrated[["umap"]])) %>% tibble(cell = colnames(integrated))

#fix cell annotations in umapCoord so that i can merge with guides table
umapCoord <- umapCoord |> 
  filter(grepl("^Trp53\\.", cell)) |> 
  mutate(cell = str_remove(cell, "^[^.]*\\."))

guides_umap <- left_join(guides, umapCoord)

```

Then we are subsetting to Trp53 enhancers only, and visualize them. 

```{r viz inert, include=TRUE}

tf_subset <- guides_umap %>% 
  filter(TF.1 == "Trp53") %>% 
  arrange(gfp_count)

trp53_single <- ggplot(tf_subset, aes(x = umap_1, y = umap_2)) + 
  geom_point(data = umapCoord, aes(x= umap_1, y = umap_2), size = 0.01, color = "lightgrey") + 
  geom_point(aes(x = umap_1, y = umap_2, color = log10(gfp_count+1)), size = 0.5) +
  theme_classic() + 
  scale_color_viridis(option = "viridis", trans = "log10") +
  facet_wrap(nrepeats~TF1.affinity+Seqname) + 
  theme(legend.position = "bottom") 

trp53_single

```

# Sequence Strength x Activity relationship

Next, we visualize the relationship between the synthetic features of the Trp53 enhancers (affinity and number of motif repeats) and the measured enhancer acitivity. For this, we first mean-aggregate all measurements per sequence. For Trp53 we have 12 different sequences.  

```{r mean aggregate}

crs_level <- tf_subset |> 
  select(Seqname, TF_combo, TF1.affinity, nrepeats, gfp_count) |> 
  dplyr::group_by(Seqname, TF_combo, TF1.affinity, nrepeats) |> 
  dplyr::summarise(mean_gfp = mean(gfp_count)) |> 
  mutate(seq_strength = TF1.affinity * nrepeats)

```

And now we can plot this sequence strength (affinity * number of repeats) x activity relationship. 

```{r plot aggregated activity}

affinity_colors <- c("#E6D2CF", "#BB9189", "#915C52", "#663329")

simple_tp53 <- ggplot(crs_level, aes(x = seq_strength, y = mean_gfp, color = as.factor(TF1.affinity))) + 
  geom_point(size = 3) + 
  labs(x = "Sequence Strength \n (Affinity x nrepeats)", y = "Mean Activity \n Log(GFP count)") + 
  scale_color_manual(values = affinity_colors) + 
  theme_minimal() +
  theme( # Remove y-axis title
    text=element_text(family="ArialMT", size=10),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank())

simple_tp53

```

Save the plot. 

```{r save linear, eval = FALSE}

setwd('~/cluster/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/thesis_figures/')
ggsave("./014_tp53_activity_design.pdf", simple_tp53, width=11, height=8, device = "pdf", units = "cm")  

```




