---
title: "Enhancer detection"
output:
  html_document:
    number_sections: yes
    toc: yes
    #code_folding: hide
    toc_depth: 4
    toc_float:
      collapsed: false
    highlight: pygments
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_knit$set(root.dir = "/users/lvelten/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/scMPRA/results/figure1")


library(tidyverse)
library(Seurat)
library(ComplexHeatmap)
library(actuar) #for zero-truncated poisson 
library(viridis)


```

In this .Rmd, we reproduce panel c and d from figure 1. For this we used
*data* from two screens: 1) PoC screen in K562 cells 2) PoC screen in
Hematopoietic Stem & Progenitor Cells (HSPCs)

We will first compute rates for enhancer detection using a maximum
likelihood approach, and afterwards plot two histograms that show
absolute detected enhancer for the the two screens.

# Enhancer Detection Rates for HSPCs

For this we need to first obtain a count table that indicates in how
many cells we detected 0 enhancers, in how many 1, ...

```{r hsc data, include=TRUE}

#Read in data 
#loading seurat object 
hsc_poc <- readRDS("/users/lvelten/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libPoC_hsc_seurat.rds")

#loading preprocessed matched_gfp_guide_enhancer_counts file
hsc_poc_guides <- readRDS("/users/lvelten/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libPoC_hsc_gfp_guide_enhancer_counts.rds")

#count the number of cells from seurat object 
no_cells_hsc_poc <- ncol(hsc_poc)

#counts in how many cells we detected 0 enhancer, 1enhancer,  2 enhancers, ...
count_hist <- c(no_cells_hsc_poc - sum(table(table(hsc_poc_guides$cell))), table(table(hsc_poc_guides$cell)))

#names vector according to the number of detected enhancers per cell 
names(count_hist) <- c("0","1", "2", "3", "4", "5")

```

Next, we write the function that determines the enhancer capture rate
given a certain:

-   lambda/MoI: the average number of molecules per cell, modeled with a
    zero-truncated Poisson

-   alpha/detection rate: the proabbility that each enhancer present in
    a cell is actually observed

This function evaluated how well different (MOI, detection) combinations
explain the observed histogram `count_hist`.

```{r hsc capture, include=TRUE}

#determine capture rates for HSCs 
hsc_presence_bc_positive_hist <- count_hist

# Candidate MOI values to test.
# In the Dixit-style notation, lambda = moi.
searchgrid_moi <- seq(0.1, 5, by = 0.5)

# Candidate detection/capture rates to test.
# alpha = probability that any one true molecule is detected.
searchgrid_detection <- seq(0.01, 0.99, by = 0.01)

# Upper cutoff for the latent true molecule count M.
# In principle M could be larger, but we truncate the sum at 20 for speed.
max_molecules <- 20

# ------------------------------------------------------------
# Log-likelihood function
#
# Inputs:
#   moi       = mean of the zero-truncated Poisson for true counts
#   guide_hist = observed histogram of counts
#   detection = per-molecule detection/capture probability
#
# Output:
#   total log-likelihood of the histogram under the model
# ------------------------------------------------------------
likelihood <- function(moi, guide_hist, detection) {

  # Compute model probability for each observed count k = 0, 1, 2, ...
  probability_k <- sapply(0:(length(guide_hist) - 1), function(k) {

    # To observe k molecules:
    # - if k = 0, the true number m can be 1, 2, 3, ...
    # - if k > 0, the true number m must be at least k
    start_m <- ifelse(k == 0, 1, k)

    # Marginalize over all possible true molecule counts m
    # P(K = k) = sum_m P(M = m) * P(K = k | M = m)
    sum(sapply(start_m:max_molecules, function(m) {

      # P(M = m): true count comes from a zero-truncated Poisson
      p_true_count <- dztpois(m, moi)

      # P(K = k | M = m): observed count comes from binomial thinning
      p_observed_given_true <- dbinom(k, size = m, prob = detection)

      # Joint contribution for this value of m
      p_true_count * p_observed_given_true
    }))
  })

  # Histogram log-likelihood:
  # if guide_hist[k+1] cells have observed count k, contribution is
  # guide_hist[k+1] * log(P(K = k))
  ll <- sum(guide_hist * log(probability_k))

  return(ll)
}

```

After defining th function, we can run it for the given parameters for
HSPCs.

```{r hsc compute and plot, include=TRUE}

combined_likelihoods = sapply(searchgrid_moi, function (x) {
  sapply(searchgrid_detection, function (y) {
    likelihood(x, hsc_presence_bc_positive_hist, y)
  })
})

T_combined_likelihoods = t(combined_likelihoods)
colnames(T_combined_likelihoods) <- searchgrid_detection
rownames(T_combined_likelihoods) <- searchgrid_moi

new <- T_combined_likelihoods 
new[new < -10000] <- -10000

hsc_guide_cap <- Heatmap(new, cluster_rows = FALSE, cluster_columns = FALSE, column_title = "GFP Detection Rate", row_title = "MOI", col=magma(100))

det_rate_enhancer = which(new==max(new), arr.ind=T)[2]
MOI_guide = rownames(which(new==max(new), arr.ind=T))

#save plot 
#pdf("./004_hsc_det_rate.pdf",width=15/2.53,height=9/2.53)
#hsc_guide_cap
#dev.off()

hsc_guide_cap

```

-   Estimated capture rate for HSPCs was: `r det_rate_enhancer`. 
  
-   Estimated MoI for HSPCs was: `r MOI_guide`.


# Enhancer Detection Rates for K562s

```{r k562 histo, include=TRUE}

k562_poc <- readRDS("/users/lvelten/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libPoC_k562_seurat.rds")

cells_to_keep <- WhichCells(k562_poc, expression = nFeature_RNA > 40 & nFeature_RNA < 200) 
k562_poc <- subset(k562_poc, cells = cells_to_keep)

#loading preprocessed matched_gfp_guide_enhancer_counts file
k562_poc_guides <- readRDS("/users/lvelten/project/SCG4SYN/Experiments/240524_scMPRA_LibBeta_HSC_iseq/screen_HSC/002_analysis/paper_figures/files_figshare/libPoC_k562_gfp_guide_enhancer_counts.rds")

#count the number of cells from seurat object 
no_cells_k562_poc <- ncol(k562_poc)

#counts in how many cells we detected 0 enhancers, 1 enhancer, ...
count_hist <- c(no_cells_k562_poc - sum(table(table(k562_poc_guides$cell))), table(table(k562_poc_guides$cell)))

#names vector according to the number of detected enhancers per cell 
names(count_hist) <- as.character(0:(length(count_hist)-1))

```

After computing the count table, we are computing again the maximum
likelihood. This time we adjust our searchgrid, since we expect a
higher MoI for the K562 cells given how we infected the cells and their
GFP+ rate.

```{r k562 likelihood, include=TRUE}

k562_presence_bc_positive_hist <- count_hist 

# Candidate MOI values to test.
searchgrid_moi <- seq(0.1, 15, by = 0.5)

# Candidate detection/capture rates to test.
searchgrid_detection <- seq(0.01, 0.99, by = 0.01)

combined_likelihoods = sapply(searchgrid_moi, function (x) {
  sapply(searchgrid_detection, function (y) {
    likelihood(x, k562_presence_bc_positive_hist, y)
  })
})

T_combined_likelihoods = t(combined_likelihoods)
colnames(T_combined_likelihoods) <- searchgrid_detection
rownames(T_combined_likelihoods) <- searchgrid_moi

new <- T_combined_likelihoods 
new[new < -30000] <- -30000

k562_guide_cap <- Heatmap(new, cluster_rows = FALSE, cluster_columns = FALSE, column_title = "GFP Detection Rate", row_title = "MOI", col=magma(100))

k562_det_rate_enhancer = which(new==max(new), arr.ind=T)[2]
k562_MOI_enhancer = rownames(which(new==max(new), arr.ind=T))

#plot

#pdf("004_k562_det_rate.pdf", width = 15 / 2.54, height = 9 / 2.54)
#k562_guide_cap
#dev.off()

k562_guide_cap

```

-   Estimated capture rate for K562s was: `r k562_det_rate_enhancer`.
  
-   Estimated MoI for K56s was: `r k562_MOI_enhancer`.

# Histograms of absolute enhancer detections per Cell

```{r histos, include=TRUE}

# build tidy data (bin 0..N-1)
df <- bind_rows(
  tibble(type = "K562",
         bin  = seq_along(k562_presence_bc_positive_hist)-1,
         count = as.numeric(k562_presence_bc_positive_hist)),
  tibble(type = "HSPC",
         bin  = seq_along(hsc_presence_bc_positive_hist)-1,
         count = as.numeric(hsc_presence_bc_positive_hist)
))

# collapse bins ≥20 into a single "20+" bin
levels_bins <- c(as.character(0:9), "10+")
df <- df %>%
  mutate(bin_cap = pmin(bin, 10L)) %>%
  group_by(type, bin_cap) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  mutate(bin = factor(ifelse(bin_cap == 10, "10+", as.character(bin_cap)),
                      levels = levels_bins))

# order the types (K562 first)
df <- df %>% mutate(type = fct_relevel(type, "K562", "HSPC"))

# plot with your template style
hist <- ggplot(df, aes(x = bin, y = count)) +
  geom_col(width = 0.9) +
  facet_wrap(~ type, ncol = 1) +
  labs(x = "Number of presence barcodes", y = "# Cells",
       title = "Barcode presence per cell") +
  theme_classic() + 
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed")
  )

hist

#for saving:
pdf("./004_presense_barcode_histograms.pdf", width = 7 / 2.54, height = 9 / 2.54)
hist
dev.off()

```

# Summary Stats for the manuscript

Last, im computing a couple of summary stats to get an overview about
cell numbers and enhancer measurements in the respective cells.

```{r summary stats, include=TRUE}

#number of enhancer postivie cells 
mult <- c(0,1,2,3,4,5)
tot_obs_hscs <- sum(hsc_presence_bc_positive_hist*mult)
enh_pos_hscs <- sum(hsc_presence_bc_positive_hist[-1])

#number of cells
no_cells_hsc_poc

#percentage of positive cells 
perc_enh_pos_hsc_cells <- enh_pos_hscs/no_cells_hsc_poc
perc_enh_pos_hsc_cells

#number of k562 observations 
mult <- seq(0, length(k562_presence_bc_positive_hist)-1)
tot_obs_k562 <- sum(k562_presence_bc_positive_hist*mult)

enh_pos_k562 <- sum(k562_presence_bc_positive_hist[-1])

no_cells_k562_poc

##percentage of positive cells 
perc_enh_pos_k562_cells <- enh_pos_k562/no_cells_k562_poc
perc_enh_pos_k562_cells

```
